Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2024-01-22 src/tex/ms.tex
[showyourwork] overleaf sync

    An interesting modification of the standard RL method is proposed by \cite{Ingaramo2014}. They show that RL can be effective in creating a merged image from multiple observations. They also observe that RL would take the best visual properties out of each image and merged them to result in a reconstruction closer to the \enquote{ground truth}.

@@

    A variation of cycle spinning specifically for the patch prior is proposed in \cite{Parameswaran2018}. Instead of a regular grid for the patches they propose to work with a randomised grid of patches, where the center of each patch is randomly shifted, while ensuring that patches still overlap by at least one pixel. In doing so they find that a smaller total number of patches can be used, while retaining the same reconstruction quality. This improves the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction is slightly faster with the same reconstruction quality. However for all results shown later in the manuscript we chose the simple cycle spinning described in the previous section, because it is the simpler method.

@@

    \section{Experiments}
    \subsection{Test Datasets}
    \label{subsec:test-datasets}
    We have designed multiple test scenarios to challenge the several deconvolution algorithms listed in Section~\ref{subsec:methods} and to evaluate and compare their performance. The test scenarios consist of mixtures of point and extended sources, with typical morphological characteristics of astrophysical sources. We arranged the sources in different spatial patterns and with different spacings between them. We include point sources superposed on extended sources \dvd{in four arrangements labelled A--D in the first row of Figure~\ref{fig:scenarios}.} Each spatial arrangement is crossed with several signal-to-noise ratios (obtained with a constant background),
    %is as well as different 
    %levels of constant background emission,
    \dvd{denoted 1--X. The resulting scenarios are denoted by concatenating the two labels, e.g., A3 or C1, see Appendix~\ref{sec:app:A} for details.}  The goal of these experiments is not to set up or devise a \enquote{perfect} deconvolution algorithm, but to help assess their relative strengths and limitations.

@@

    The ground truth and corresponding expected counts for all combinations of source and instrument scenarios are shown in Figure~\ref{fig:scenarios}. The details of the locations and brightness of each component is listed in Table~\ref{tab:scenarios}. 
    %In addition we also 
    We assume a constant background with varying levels of $\lambda_{Bkg}= 0.001, 0.01$ and \qty[mode = text]{0.1}{counts/pixel}.

@@

        \item[\bf M1.] \textbf{Pylira}: we implement the LIRA method \citep{Esch2004} via the Pylira Python wrapper \citep{Donath2022}. We rely on the default prior definition and keep the background norm fixed. In total we run the MCMC sampler for 5000 iterations and discard the first 500 iterations as a \enquote{burn-in} period to allow the MCMC chain to converge. We compute the mean of the posterior distribution ignoring the burn-in period.

@@

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of the different deconvolution methods and prior assumptions for selected source scenarios, as described in Section~\ref{subsec:test-datasets}. The simulations depicted used a fixed background level of $\lambda_{Bkg} = \qty[mode = math]{0.01}{cts/pix}$, a uniform exposure, and a Gaussian PSF of width $\sigma_{PSF} = \qty[mode = math]{2}{pix}$, corresponding to the \enquote{Chandra} instrument scenario. The leftmost column shows the simulated counts and the second column shows the underlying ground truth. The remaining columns show the reconstructed flux for each method and prior assumptions for \jolideconospace. To enhance weak structures the images use an $\arcsinh$ stretching with a scale parameter of $a=0.02$. The stretching is the same for all images. The methods are described in detail in Section~\ref{subsec:methods}. A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
            \script{comparison-scenarios.py}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of the five methods (and their differing prior assumptions) on the different source scenarios. Figure \ref{fig:comparison-scenarios} shows the results for the selected subset of scenarios, i.e., on A1, B3, C3 and D4.  \dvd{What do the numbers refer to?}

@@

    \section{Application Examples}
    %
    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-chandra.pdf}
            \caption{
               \jolideco result for \chandra observations of SNR \textit{1E~0102.2–7219}. The bottom left image shows the summed counts from 25 observations. The bottom right image  shows the the \jolideco deconvolution. The flux is given as a counts rate relative to obsID 1308. The panels \textit{Zoom A}, \textit{Zoom B} and \textit{Zoom C} show a zoomed-in version of three example regions in the SNR, marked with the white rectangles in the large images. The colormap and scale used in the zoom insets is the same as used for the larger images.
            }
            \label{fig:example-chandra}
            \script{example-chandra.py}
        \end{centering}
    \end{figure*}
    %
    \subsection{Deep \chandra Observation}
    \label{sec:chandra-example}
    \subsubsection{Image Example}
    First we apply \jolideco to a set of short-exposure \chandra observations of the SNR \textit{1E~0102.2–7219} and combine them into a single long-exposure image using the patch prior learned from JWST data. In total, we chose 25 observations\footnote{Same list of observation IDs as for \url{https://chandra.si.edu/photo/2009/e0102/}} taken with  the ACIS-S instrument and no grating.

    We reduced the data using \texttt{ciao} v4.15 and a custom \texttt{snakemake} workflow (see also Section~\ref{sec:reproducibility} on reproducibility). For the counts image we selected an energy range from \qty[mode = text]{0.5}{keV} to  \qty[mode = text]{7}{keV} and a pixel size of $0.25$ of the native pixel size of \chandra. This corresponds to an absolute  pixel size of $\approx$~\qty[mode = text]{0.125}{$\arcsec$/pixel} or a sampling of $\approx2$ pixels for the $38\%$ containment radius of Chandra's PSF. \chandra's instrument response varies with the observation offset and energy \citep{ChandraPOG2022}, so we simulated the PSF for each observation independently using the \texttt{marx} tool \citep{Davis2012} and the \texttt{simulate\_psf} script. As a reference spectrum for the PSF simulation we assumed the model derived by \cite{Plucinsky2017}. We converted the spectral script from XSPEC to sherpa using the helper script \texttt{convert\_xspec\_script}. As reported in \cite{Xi2019} the effect of pileup, even for the brightest regions is small ($<5\%$), so we did not include the effect of pileup in our PSF simulations.

@@

    \subsubsection{Flux Profile}
    To quantify the improvement in angular resolution of the reconstructed image over the data we compute a flux profile on the \jolideco reconstruction of E0102. We chose a region perpendicular to the filament shown in Zoom~A of Figure~\ref{fig:example-chandra}. We compute errors via the bootstrap, with resampled data obtained by sampling the event lists with replacement for each individual observation. For each resampled data set we run \jolideco and compute a flux profile on the reconstructed image. In this way we obtain a distribution of profiles, which we can use to compute an error band. For comparison we also compute the profiles and associated errors directly on the stacked bootstrapped counts (without processing or fitting). The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

    We compute the full-width at half-max (FWHM) $\omega$ of each of the profiles, by fitting a Gaussian shape with a linear gradient as a baseline for both the original counts and the \jolideconospace-deconvolved intensities. In both cases we compute the mean of $\omega$. We find that the deconvolved profile is significantly narrower. For the counts profile, $\omega_C=0.77 \pm 0.22$~arcsec, where the error bar is obtained by computing the standard deviations over the bootstrap sample; and the deconvolved intensity profile has $\omega_J=0.52 \pm 0.03$~arcsec.

    Assuming that the PSF and the shape of profile are approximately Gaussian we expect
    the width of the counts profile, the PSF, and the \jolideconospace-reconstructed profile to be related as:
    %
    \begin{equation}
        \omega_{J} = \sqrt{\omega_C^2 - \omega_{\mathrm{PSF}}^2}
    \end{equation}
    %
    With $\omega_{\mathrm{PSF}}\approx0.6$~arcsec\footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}, we expect a FWHM after deconvolution of $\omega_J\approx0.48$~arcsec. This is roughly consistent with the values we measured for the profile.

@@

    \subsection{\fermi Event Types}

     We apply \jolideco  to data from the Fermi Large Area Telescope (LAT) \citep{Atwood2009}. The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose  a Galactic source with complex morphology: the supernova remnant Vela Junior. We use 14 years (from 2008 August 4 to 2023 August 4) of \enquote{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. This is consistent with the data selection range in \cite{Ackermann2017}. To preserve as much information as possible on the angular resolution of the instrument, we split the data into the four event types \enquote{PSF0}, \enquote{PSF1}, \enquote{PSF2} and \enquote{PSF3}, as defined by the \fermi \enquote{Pass8} data release\footnote{\url{https://fermi.gsfc.nasa.gov/ssc/data/analysis/documentation/Pass8_usage.html}}. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5 of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \citep{Fermitools2019}.

@@

    The choice of the prior becomes more important at low signal noise ratios. The right choice can yield superior reconstruction quality then any other method we have compared to. The advantage of the GMM patch prior is its expressiveness and that it is adapted to the data in the optimization process. However if the GMM patch prior is learned on unsuited data or data that not diverse enough it can contain a strong bias and there is the danger of hallucinating features. Testing of systematic uncertainties therefore becomes crucial.

>>>>

    An interesting modification of the standard RL method has been proposed by \cite{Ingaramo2014}. They show that RL could be effective in creating a merged image from multiple observations. They also observed that RL would take the best visual properties out of each image and merged them to result in a reconstruction closer to the \enquote{ground truth}.

@@

    A variation of cycle spinning specifically for the patch prior is proposed in \cite{Parameswaran2018}. Instead of a regular grid for the patches they propose to work with a randomised grid of patches, where the center of each patch is randomly shifted, while ensuring that patches still overlap by at least one pixel. In doing so they find that a smaller total number of patches can be used, while retaining the same reconstruction quality. This improves the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction is slightly faster with the same reconstruction quality. However for all results shown later in the manuscript we chose the simpler cycle spinning described in the previous section.

@@

    \section{Experiments}
    \subsection{Test Datasets}
    \label{subsec:test-datasets}
    We have designed multiple test scenarios to challenge the several deconvolution algorithms listed in Section~\ref{subsec:methods} and to evaluate and compare their performance. The test scenarios consist of mixtures of point and extended sources, with typical morphological characteristics of astrophysical sources. We arranged the sources in different spatial patterns and with different spacings between them. We include point sources superposed on extended sources in four arrangements labelled A--D in the first row of Figure~\ref{fig:scenarios}. Each spatial arrangement is crossed with several signal-to-noise ratios (obtained with a constant background),
    denoted \texttt{XN}. The resulting scenarios are denoted by concatenating the two labels, e.g., A3 or C1, see Appendix~\ref{sec:app:A} for details.  The goal of these experiments is not to set up or devise a \enquote{perfect} deconvolution algorithm, but to help assess their relative strengths and limitations.

@@

    The ground truth and corresponding expected counts for all combinations of source and instrument scenarios are shown in Figure~\ref{fig:scenarios}. The details of the locations and brightness of each component is listed in Table~\ref{tab:scenarios}. 
    We assume a constant background with varying levels of $\lambda_{Bkg}= 0.001, 0.01$ and \qty[mode = text]{0.1}{counts/pixel}.

@@

        \item[\bf M1.]: we implement the LIRA method \citep{Esch2004} via the Pylira Python wrapper \citep{Donath2022}. We rely on the default prior definition and keep the background norm fixed. In total we run the MCMC sampler for 5000 iterations and discard the first 500 iterations as a \enquote{burn-in} period to allow the MCMC chain to converge. We compute the mean of the posterior distribution ignoring the burn-in period.

@@

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of the different deconvolution methods and prior assumptions for selected source scenarios, as described in Section~\ref{subsec:test-datasets}. The simulations depicted used a fixed background level of $\lambda_{Bkg} = \qty[mode = math]{0.01}{cts/pix}$, a uniform exposure, and a Gaussian PSF of width $\sigma_{PSF} = \qty[mode = math]{2}{pix}$, corresponding to the \enquote{Chandra} instrument scenario. The leftmost column shows the simulated counts and the second column shows the underlying ground truth. The remaining columns show the reconstructed flux for each method and prior assumptions for \jolideconospace. To enhance weak structures the images use an $\arcsinh$ stretching with a scale parameter of $a=0.02$. The stretching is the same for all images. The methods are described in detail in Section~\ref{subsec:methods}. A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
            \script{comparison-scenarios.py}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of the five methods and their differing prior assumptions on the different source scenarios. Figure \ref{fig:comparison-scenarios} shows the results for the selected subset of scenarios, i.e., on A1, B3, C3 and D4.

@@

    \section{Application Examples}
    %
    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-chandra.pdf}
            \caption{
               \jolideco result for \chandra observations of SNR \textit{1E~0102.2–7219}. The bottom left image shows the summed counts from 25 observations. The bottom right image  shows the the \jolideco deconvolution. The flux is given as a counts rate relative to obsID 1308. The panels \textit{Zoom A}, \textit{Zoom B} and \textit{Zoom C} show a zoomed-in version of three example regions in the SNR, marked with the white rectangles in the large images. The colormap and scale used in the zoom insets is the same as used for the larger images.
            }
            \label{fig:example-chandra}
            \script{example-chandra.py}
        \end{centering}
    \end{figure*}
    %
    \subsection{Deep \chandra Observation}
    \label{sec:chandra-example}
    \subsubsection{Image Example}
    First we apply \jolideco to a set of short-exposure \chandra observations of the SNR \textit{1E~0102.2–7219} and combine them into a single long-exposure image using the patch prior learned from JWST data. In total, we chose 25 observations taken with  the ACIS-S instrument and no grating. We chose newer data, taken after 2006, to limit the effect of pile up.

    We reduced the data using \texttt{ciao} v4.15 and a custom \texttt{snakemake} workflow (see also Section~\ref{sec:reproducibility} on reproducibility). For the counts image we selected an energy range from \qty[mode = text]{0.5}{keV} to  \qty[mode = text]{7}{keV} and a pixel size of $0.25$ of the native pixel size of \chandra. This corresponds to an absolute  pixel size of $\approx$~\qty[mode = text]{0.125}{$\arcsec$/pixel} or a sampling of $\approx2$ pixels for the $38\%$ containment radius of Chandra's PSF. \chandra's instrument response varies with the observation offset and energy \citep{ChandraPOG2022}, so we simulated the PSF for each observation independently using the \texttt{marx} tool \citep{Davis2012} and the \texttt{simulate\_psf} script. As a reference spectrum for the PSF simulation we assumed the model derived by \cite{Plucinsky2017}. We converted the spectral script from XSPEC to sherpa using the helper script \texttt{convert\_xspec\_script}.

@@

    \subsubsection{Flux Profile}
    To quantify the improvement in angular resolution of the reconstructed image over the data we compute a flux profile on the \jolideco reconstruction of E0102. We chose a region perpendicular to the filament shown in Zoom~A of Figure~\ref{fig:example-chandra}. We compute errors via the bootstrap, with resampled data obtained by sampling the event lists with replacement for each individual observation. For each resampled data set we run \jolideco and compute a flux profile on the reconstructed image. In this way we obtain a distribution of profiles, which we use to compute an error band. For comparison we also compute the profiles and associated errors directly on the stacked bootstrapped counts (without processing or fitting). The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

    We compute the full-width at half-max (FWHM) $\omega$ of each of the profiles, by fitting a Gaussian shape with a linear gradient as a baseline for both the original counts and the \jolideconospace-deconvolved intensities. In both cases we compute the mean of $\omega$. We find that the deconvolved profile is significantly narrower. For the counts profile, $\omega_C=0.78 \pm 0.02$~arcsec, where the error bar is obtained by computing the standard deviations over the bootstrap sample; and the deconvolved intensity profile has $\omega_J=0.58 \pm 0.02$~arcsec.

    Assuming that the PSF and the shape of profile are approximately Gaussian we expect the width of the counts profile, the PSF, and the \jolideconospace-reconstructed profile to be related as:
    %
    \begin{equation}
        \omega_{J} = \sqrt{\omega_C^2 - \omega_{\mathrm{PSF}}^2}
    \end{equation}
    %
    With $\omega_{\mathrm{PSF}}\approx0.6$~arcsec\footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}, we expect a FWHM after deconvolution of $\omega_J\approx0.5$~arcsec. This is roughly consistent with the values we measured for the profile.

@@

    \subsection{\fermi Event Types}
     As a second application example, we apply \jolideco  to data from the Fermi Large Area Telescope (LAT) \citep{Atwood2009}. The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose  a Galactic source with complex morphology: the supernova remnant Vela Junior. We use 14 years (from 2008 August 4 to 2023 August 4) of \enquote{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. This is consistent with the data selection range in \cite{Ackermann2017}. To preserve as much information as possible on the angular resolution of the instrument, we split the data into the four event types \enquote{PSF0}, \enquote{PSF1}, \enquote{PSF2} and \enquote{PSF3}, as defined by the \fermi \enquote{Pass8} data release\footnote{\url{https://fermi.gsfc.nasa.gov/ssc/data/analysis/documentation/Pass8_usage.html}}. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5 of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \citep{Fermitools2019}.

@@

    % The choice of the prior becomes more important at low signal noise ratios. The right choice can yield superior reconstruction quality then any other method we have compared to. The advantage of the GMM patch prior is its expressiveness and that it is adapted to the data in the optimization process. However if the GMM patch prior is learned on unsuited data or data that not diverse enough it can contain a strong bias and there is the danger of hallucinating features. Testing of systematic uncertainties therefore becomes crucial.
